Preferences for the commodification of pensions in Chile: the rol of intergenerational social mobility

### 2025-11-30 paper/paper.tex
Up antecedentes

\setstretch{1.5}
This document was last modified at 2025-11-21 21:08:43 and it was last
rendered at 2025-11-21 21:08:43.

@@

What is the legitimate scope of market inequality in the eyes of the
public? Since the early 1980s, many countries have experienced a
rollback of universal welfare programs and a shift toward the
privatisation and commodification of public goods and social services
(\citeproc{ref-gingrich_making_2011}{Gingrich, 2011};
\citeproc{ref-streeck_how_2016}{Streeck, 2016}). In Latin America, these
reforms extended market logic into domains of social reproduction long
governed by the state, shrinking public provision and expanding private
actors in core welfare sectors (\citeproc{ref-ferre_welfare_2023}{Ferre,
2023}). From a moral-economy perspective, this diffusion of market rules
has reconfigured the normative order: through policy feedback, welfare
institutions embed market-based criteria of ``fair'' allocation and
shape how citizens understand deservingness and the balance between
state and market (\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
\citeproc{ref-svallforsMoralEconomyClass2006a}{Svallfors, 2006}).
Against this backdrop, growing scholarship examines \emph{market justice
preferences}---the extent to which citizens regard market-based criteria
as a fair basis for access to essential services such as healthcare,
education and pensions
(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
2025}; \citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). These
orientations matter because they help legitimise economic inequality by
portraying stratified outcomes as a fair result of individual
responsibility and limited state intervention.

This article examines public support for the commodification of pension
welfare---specifically, whether people consider it fair that access to
better pensions depends on income or contributions. Pension
privatisation lies at the core of Latin America's trajectory of
commodification (\citeproc{ref-huber_political_2000}{Huber \& Stephens,
2000}). As many countries expanded mandatory or voluntary
individual-capitalisation schemes---most radically in Chile---pensions
became organised around contribution-based entitlements and market
performance (\citeproc{ref-arenas_sistemas_2019}{Arenas, 2019};
\citeproc{ref-verbic_political_2019}{Verbič \& Spruk, 2019}). This shift
reshapes how pension institutions allocate old-age risk, tying benefit
levels to lifetime labour-market trajectories and increasing exposure to
poverty in retirement
(\citeproc{ref-madero-cabib_private_2019}{Madero-Cabib et al., 2019};
\citeproc{ref-oecd_pensions_2023}{OECD, 2023}). It also transforms the
justice principles citizens associate with old-age security: they must
judge whether higher benefits should follow market-derived criteria or
solidaristic principles
(\citeproc{ref-arrizabalomontoro_debate_2019}{Arrizabalo Montoro et al.,
2019}; \citeproc{ref-borzutzky_pension_2012}{Borzutzky, 2012}). Although
we know something about what people consider a ``just'' monetary pension
(\citeproc{ref-castillo_deserving_2019}{Castillo, Olivos, et al., 2019})
and about support for public versus private provision
(\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020};
\citeproc{ref-jaime-castillo_public_2013}{Jaime-Castillo, 2013}), we
still lack systematic evidence on how citizens assess the fairness of
market-dependent access to pensions.

Research shows that support for market-based access to welfare services
is strongly stratified: individuals with higher income, education or
class position are more likely to endorse market-based criteria than
disadvantaged groups (\citeproc{ref-busemeyer_skills_2014}{Busemeyer,
2014}; \citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). Evidence for
pensions points in the same direction: higher income and greater
financial gains in funded schemes correlate with stronger support for
market-oriented pension policies
(\citeproc{ref-kerner_pension_2020}{Kerner, 2020}). Yet we know very
little about how movement within the class structure shapes preferences
for pension commodification. This gap is especially salient in Latin
America, where mobility has unfolded under high inequality and deeply
privatised welfare regimes
(\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
2021}). Social origins and destinations anchor material interests and
justice principles
(\citeproc{ref-alesina_intergenerational_2018}{Alesina et al., 2018};
\citeproc{ref-gugushvili_subjective_2017}{Gugushvili, 2017};
\citeproc{ref-langsaether_explaining_2022}{Langsæther et al., 2022}),
while movement between these positions exposes individuals to new risks,
resources and normative environments that reshape fairness judgements
(\citeproc{ref-gugushvili_trends_2014}{Gugushvili, 2014};
\citeproc{ref-helgason_longterm_2023}{Helgason \& Rehm, 2023}). Social
mobility can thus foster market-justice orientations, helping explain
how inequalities in access to welfare services become normatively
sustained (\citeproc{ref-mau_inequality_2015}{Mau, 2015}).

Existing accounts identify several mechanisms through which
intergenerational mobility may shape distributive-justice preferences
but converge on two channels particularly relevant for market-based
welfare: changes in material stakes and shifts in normative frames
(\citeproc{ref-ares_changing_2020}{Ares, 2020};
\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
\citeproc{ref-helgason_longterm_2023}{Helgason \& Rehm, 2023};
\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
Marqués-Perales, 2019}). Recent research shows that meritocratic
beliefs---seeing that effort and ability are rewarded---are strongly
associated with support for market allocation
(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
2025}). Perceived upward mobility reinforces meritocratic
interpretations of success, whereas downward mobility is more closely
linked to structural explanations of inequality without necessarily
eroding meritocratic convictions (\citeproc{ref-mijs_belief_2022}{Mijs
et al., 2022}). Thus, meritocracy and mobility operate on partly
separate tracks yet intersect in systematic ways. I treat meritocratic
beliefs as a post-treatment moderator---a diagnostic lens on potential
mechanisms. If mobility partly shapes preferences by strengthening
merit-based understandings of fairness and altering material stakes, its
effects should be most visible---and most pro-market---among individuals
who already interpret social outcomes through meritocratic lenses.

This article pursues two aims: first, to assess whether
intergenerational occupational mobility causally affects support for the
commodification of pensions; and second, to examine whether this effect
varies by meritocratic beliefs as a diagnostic of possible transmission
channels. I argue that social mobility operates through two main
pathways. Upward mobility increases earnings, expected returns in a
funded system and perceived economic security, making contribution-based
access more appealing. At the same time, exposure to higher-class
networks reinforces individualistic and effort-based understandings of
fairness. Downward mobility tends to heighten economic insecurity,
dependence on public protection and attachment to external attributions
of poverty, reducing support for market allocation. Meritocratic beliefs
are conceptualised not as mediators but as moderators. Among individuals
with stronger meritocratic beliefs, upward mobility should heighten
support for market-based access. In contrast, downward mobility should
weaken---or even reverse---its usual de-commodifying pattern, as
meritocratic interpretations of effort lead individuals to view market
outcomes as fair regardless of direction. The paper contributes to
research on market justice by identifying the causal effects of mobility
and by using meritocracy to probe whether these effects align with a
meritocratic channel.

The study focuses on Chile, an instructive case for examining
preferences for pension commodification. Despite sustained economic
growth, Chile combines high and persistent inequality with short-range
upward mobility and substantial barriers to higher-class positions
(\citeproc{ref-flores_top_2020}{Flores et al., 2020};
\citeproc{ref-salgado_uplifting_2025}{Salgado et al., 2025};
\citeproc{ref-torche_intergenerational_2014}{Torche, 2014}). It was the
first country worldwide to fully replace a public pay-as-you-go system
with a mandatory, privately administered defined-contribution scheme,
while the state has progressively expanded a segmented solidarity pillar
to compensate for insufficient benefits
(\citeproc{ref-boccardo_30_2020}{Boccardo, 2020};
\citeproc{ref-madariaga_three_2020}{Madariaga, 2020};
\citeproc{ref-solimano_rise_2021}{Solimano, 2021}). In parallel---and
despite waves of protest from 2006 to 2019 against the private pension
system (\citeproc{ref-somma_no_2021}{Somma et al., 2021})---Chilean
subjectivities have been increasingly shaped by neoliberal discourses
and market logics, influencing attitudes toward pension distribution and
inequality (\citeproc{ref-araujo_desafios_2012}{Araujo \& Martuccelli,
2012}; \citeproc{ref-galvez_con_2023}{Gálvez, 2023}). The combination of
profound inequality, market-driven old-age security and contested
legitimacy makes Chile a critical case for analysing how mobility and
meritocratic beliefs structure support for pension commodification.

@@

\subsection{Preferences for market-based
welfare}\label{preferences-for-market-based-welfare}

Beyond state's capacity to reallocate resources from the advantaged to
the vulnerable, market institutions also play a central role in
distributing socially valuable goods and rewards
(\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}). As Polanyi
(\citeproc{ref-polanyi_great_1975}{1975}) observed, economic integration
in capitalist societies is primarily organized through market exchange,
governed by a self-regulating price system embedded in institutional
frameworks. These institutions are not mere aggregates of individual
behavior, but social realities endowed with rules, mechanisms, and
normative meanings that shape everyday thinking
(\citeproc{ref-immergut_theoretical_1998}{Immergut, 1998};
\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019}). In this sense,
the economic order is mirrored in a moral economy: collectively shared
norms and beliefs about justice in distribution, embedded and reinforced
through institutions (\citeproc{ref-mau_inequality_2015}{Mau, 2015};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). While most
research from this perspective has focused on welfare institutions,
recent scholarship has brought the market back into focus as a site of
distributive justice beliefs and institutional responses to inequality
(\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}). In many
countries, privatization and commodification have expanded market logic
into core areas of social reproduction, such as healthcare, education,
pensions, and social security, deepening inequality in access to these
services (\citeproc{ref-ferre_welfare_2023}{Ferre, 2023};
\citeproc{ref-gingrich_making_2011}{Gingrich, 2011}). Yet public support
for market-based welfare provision has grown, even in traditional
welfare states, particularly among higher-income groups who view private
alternatives as more efficient or higher in quality
(\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020}).
This shift calls for broader inquiry into how institutions like markets
structure access to resources and legitimize inequalities
(\citeproc{ref-mau_inequality_2015}{Mau, 2015};
\citeproc{ref-satz_por_2019}{Satz, 2019}).

The legitimacy of market-based inequalities is closely tied to beliefs
about distributive justice grounded in market principles. Beyond
stratification structures and socioeconomic conditions, the study of
inequality also focuses on individuals' beliefs about the origins of
inequality, the normative frameworks that sustain those beliefs, the
mechanisms that shape them, and their implications for attitudes and
behavior (\citeproc{ref-kluegel_beliefs_1981}{Kluegel \& Smith, 1981}).
In this regard, empirical research on distributive justice focuses on
individuals' conceptions of how goods and rewards \emph{should} be
distributed in society (\citeproc{ref-jasso_distributive_2016}{Jasso et
al., 2016}). This line of inquiry allows for examining the extent to
which economic inequality is perceived as just, or in a certain sense,
legitimate (\citeproc{ref-castillo_legitimacy_2011}{Castillo, 2011}).
Most of the research on this field explores the legitimacy of wage
inequality, specially salary gaps between jobs at opposite ends of the
occupational hierarchy
(\citeproc{ref-castillo_legitimacy_2011}{Castillo, 2011};
\citeproc{ref-jasso_justice_1978}{Jasso, 1978};
\citeproc{ref-jasso_gender_1999}{Jasso \& Wegener, 1999}). Recently,
others examines how individuals justify market-generated inequalities in
access to core social services such as healthcare, education, pensions
or social security. Here, legitimacy stems from the belief that access
to these goods should follow market-based criteria
(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). In such views, these
services are treated as legitimate commodities; goods that can be
traded, priced, and evaluated through market logic
(\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020}).

Market justice preferences refer to normative beliefs that legitimize
the idea that access to core social services---such as healthcare,
education, or pensions---should be determined by market-based criteria.
Following Janmaat's (\citeproc{ref-janmaat_subjective_2013}{2013, p.
359}) distinction, these preferences fall under the category of
``beliefs,'' understood as normative ideas about what inequality should
look like, as opposed to ``perceptions,'' which refer to subjective
evaluations of existing inequality. Market justice preferences reflect
the view that access to these services should depend on individuals'
ability to pay, thus justifying inequalities generated by market
mechanisms (\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al.,
1999}; \citeproc{ref-lindh_public_2015}{Lindh, 2015}). The concept draws
on Lane's (\citeproc{ref-lane_market_1986}{1986}) classic contrast
between market justice and political justice: the former is grounded in
the idea of earned deserts---where rewards reflect effort, productivity,
and skill---while the latter prioritizes need and equality, typically
expressed in welfare state policies. Lane argued that markets and states
differ in purpose (efficiency vs.~need), logic (individual
vs.~collective), and fairness criteria (merit vs.~equality). Market
justice assumes that markets are neutral, self-regulating systems in
which fair procedures yield outcomes proportional to merit. Inequality,
from this perspective, is not only expected but legitimate---as long as
it arises from fair competition. In this way, market justice offers a
moral lens through which individuals can view the commodification and
stratified access to social services as fair and justified
(\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al., 1999};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}).

The empirical study of market justice preferences has employed various
strategies to capture how individuals assess inequalities arising from
market allocation. A common approach evaluates whether people find it
fair that access to essential social services---such as healthcare,
education, or pensions---depends on income. This builds on foundational
work by Kluegel and Smith
(\citeproc{ref-kluegel_legitimation_1999}{1999}), who examined normative
justifications for capitalist systems. Subsequent studies have extended
this logic to a wider set of welfare goods. For example, Immergut and
Schneider (\citeproc{ref-immergut_it_2020}{2020}) and von dem Knesebeck
et al. (\citeproc{ref-vondemknesebeck_are_2016}{2016}) explore whether
respondents believe it is just that higher-income individuals receive
better healthcare. Similarly, Lee and Stacey
(\citeproc{ref-lee_fairness_2023}{2023}) and Castillo, Iturra, et al.
(\citeproc{ref-castillo_changes_2025}{2025}) apply this framework to
education. Complementing these efforts, more recent research has
introduced composite indicators to capture broader orientations toward
market-based allocation. Lindh (\citeproc{ref-lindh_public_2015}{2015}),
for instance, constructs an index averaging support for income-based
access to healthcare and education in comparative perspective, while
Castillo et al. (\citeproc{ref-castillo_socialization_2024}{2024})
propose a single-item measure covering health, education, and pensions
in Chile. These instruments seek to gauge to what extent individuals
view market-generated inequalities as legitimate. Taken together, they
capture two core dimensions of market distribution: the role of economic
resources as a key determinant of outcomes, and the framing of services
as tradable commodities that can be bought and sold according to ability
to pay (\citeproc{ref-lindh_public_2015}{Lindh, 2015}).

Comparative empirical research has identified several individual-level
factors that influence support for market justice. Individuals in more
advantaged socioeconomic positions---those with higher income,
education, and occupational status---are consistently more likely to
endorse market-based distributive principles
(\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). For example,
Lindh (\citeproc{ref-lindh_public_2015}{2015}) finds that individuals
from the service class are more likely to support market-based access to
healthcare and education than skilled and unskilled workers across 17
relatively affluent countries. In a comparative analysis, Svallfors
(\citeproc{ref-svallfors_political_2007}{2007}) observes that this
expected class pattern appears clearly only in Sweden, where support for
private education and healthcare varies systematically by class.
Busemeyer (\citeproc{ref-busemeyer_skills_2014}{2014}) similarly shows
that support for private education is stronger among high-income groups,
while Immergut and Schneider (\citeproc{ref-immergut_it_2020}{2020}) and
von dem Knesebeck et al. (\citeproc{ref-vondemknesebeck_are_2016}{2016})
report comparable findings for healthcare, suggesting that wealthier
individuals perceive private provision as a means of maintaining
relative advantage. In Chile, Otero and Mendoza
(\citeproc{ref-otero_power_2024}{2024}) show that individuals with
higher income and university education express stronger support for
market allocation in healthcare, education, and pensions. Beliefs about
inequality and political orientation also matter. Castillo et al.
(\citeproc{ref-castillo_socialization_2024}{2024}) and Castillo,
Laffert, et al. (\citeproc{ref-castillo_perceptions_2025}{2025}) show
that individuals who have strong meritocratic beliefs are more likely to
support market-based distribution in Chile, while Lee and Stacy
(\citeproc{ref-lee_fairness_2023}{2023}) in Australia suggest that these
preferences are also greater as people lean toward economic
conservatism. In this sense, market justice preferences are shaped by
the interplay between structural position and normative reasoning.

>>>>

\setstretch{1.5}
This document was last modified at 2025-11-29 23:44:43 and it was last
rendered at 2025-11-29 23:44:43.

@@

What is the legitimate scope of market inequality in the eyes of the
public? Since the early 1980s, many countries have experienced a
rollback of universal welfare programs and a shift toward the
privatisation and commodification of public goods and social services
(\citeproc{ref-gingrich_making_2011}{Gingrich, 2011};
\citeproc{ref-streeck_how_2016}{Streeck, 2016}). In Latin America, these
reforms extended market logic into domains of social reproduction long
governed by the state, shrinking public provision and expanding private
actors in core welfare sectors (\citeproc{ref-ferre_welfare_2023}{Ferre,
2023}). From a moral-economy perspective, this diffusion of market rules
has reconfigured the normative order: through policy feedback, welfare
institutions embed market-based criteria of ``fair'' allocation and
shape how citizens understand deservingness and the balance between
state and market (\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
\citeproc{ref-svallforsMoralEconomyClass2006a}{Svallfors, 2006}).
Against this backdrop, growing scholarship examines \emph{market justice
preferences}---the extent to which citizens regard market-based criteria
as a fair basis for the allocation of essential services such as
healthcare, education and pensions
(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
2025}; \citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). These
orientations matter because they legitimize unequal outcomes as the
product of individual responsibility and promote the conception of basic
social services as commodities.

This article examines public support for the commodification of pension
welfare---specifically, whether people consider it fair that access to
better pensions depends on income or contributions. Pension
privatisation lies at the core of Latin America's trajectory of
commodification (\citeproc{ref-huber_political_2000}{Huber \& Stephens,
2000}). As many countries expanded mandatory or voluntary
individual-capitalisation schemes---most radically in Chile---pensions
became organised around contribution-based entitlements and market
performance (\citeproc{ref-arenas_sistemas_2019}{Arenas, 2019};
\citeproc{ref-verbic_political_2019}{Verbič \& Spruk, 2019}). This shift
reshapes not only how pension institutions allocate old-age risk and tie
benefit levels to lifetime labour-market trajectories
(\citeproc{ref-madero-cabib_private_2019}{Madero-Cabib et al., 2019};
\citeproc{ref-oecd_pensions_2023}{OECD, 2023}), but also transforms the
justice principles citizens associate with old-age security: they must
judge whether higher benefits should follow market-derived criteria or
solidaristic principles
(\citeproc{ref-borzutzky_pension_2012}{Borzutzky, 2012};
\citeproc{ref-mau_inequality_2015}{Mau, 2015}). Although we know
something about what people consider a ``just'' monetary pension
(\citeproc{ref-castillo_deserving_2019}{Castillo, Olivos, et al., 2019})
and about support for public versus private provision
(\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020};
\citeproc{ref-jaime-castillo_public_2013}{Jaime-Castillo, 2013}), we
still lack systematic evidence on how citizens assess the fairness of
market-dependent access to pensions.

Research shows that support for market-based access to welfare services
is strongly stratified: people with higher socioeconomic status are more
likely to support market-based criteria than disadvantaged groups
(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). Evidence for
pensions points in the same direction: higher income and greater
financial gains in funded schemes correlate with stronger support for
market-oriented pension policies
(\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020};
\citeproc{ref-kerner_pension_2020}{Kerner, 2020}). Yet we know very
little about how movement within the class structure shapes preferences
for pension commodification. This gap is especially salient in Latin
America, where mobility has unfolded under high inequality and deeply
privatised welfare regimes
(\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
2021}). Social origins and destinations anchor material interests and
justice principles
(\citeproc{ref-alesina_intergenerational_2018}{Alesina et al., 2018};
\citeproc{ref-gugushvili_subjective_2017}{Gugushvili, 2017};
\citeproc{ref-langsaether_explaining_2022}{Langsæther et al., 2022}),
while movement between these positions exposes individuals to new risks,
resources and normative environments that reshape fairness judgements
(\citeproc{ref-gugushvili_trends_2014}{Gugushvili, 2014};
\citeproc{ref-helgason_longterm_2023}{Helgason \& Rehm, 2023}). Social
mobility can thus foster market-justice orientations, helping explain
how inequalities in access to welfare services become normatively
sustained (\citeproc{ref-mau_inequality_2015}{Mau, 2015}).

Existing accounts identify several mechanisms through which
intergenerational mobility may shape distributive justice preferences
but converge on two channels especially relevant for market-based
welfare: changes in material interests and shifts in normative frames
(\citeproc{ref-ares_changing_2020}{Ares, 2020};
\citeproc{ref-gugushvili_intergenerational_2016c}{Gugushvili, 2016a};
\citeproc{ref-helgason_longterm_2023}{Helgason \& Rehm, 2023};
\citeproc{ref-jaime-castillo_social_2019}{Jaime-Castillo \&
Marqués-Perales, 2019}). Mobility alters individuals' material stakes by
changing their occupational status, income, and risk exposure. At the
same time, it reshapes their fairness judgments. Empirically, upward
mobility is often associated with stronger meritocratic interpretations
of success and fairness, whereas downward mobility is more closely
linked to structural explanations of inequality without necessarily
eroding meritocratic convictions (\citeproc{ref-bucca_merit_2016}{Bucca,
2016}; \citeproc{ref-deng_its_2025}{Deng \& Wang, 2025};
\citeproc{ref-mijs_belief_2022}{Mijs et al., 2022}). Regarding
market-based preferences, recent research shows that meritocratic
beliefs -- the conviction that effort and ability are rewarded -- are
strongly associated with support for market allocation
(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
2025}). In such contexts, the upwardly mobile are especially likely to
view markets as just distributors, since their own trajectories appear
to confirm that markets appropriately reward effort; those who
experience downward mobility may still interpret outcomes through a
meritocratic lens, often reading losses as personal failure rather than
as evidence against market fairness. Thus, mobility and meritocracy
evolve on partly separate tracks yet intersect systematically
(\citeproc{ref-mau_inequality_2015}{Mau, 2015}).

Against this backdrop, the central aim of this article is to assess
whether intergenerational occupational mobility causally affects support
for the commodification of pensions. A second, more exploratory aim is
to examine whether this effect varies by meritocratic beliefs, using
this heterogeneity as an indication of possible underlying mechanisms. I
argue that social mobility operates through two main pathways
(\citeproc{ref-mau_inequality_2015}{Mau, 2015}). Upward mobility
increases earnings, expected returns in a funded system, and perceived
economic security, making contribution-based access more appealing.
Simultaneously, upward moves reinforce individualistic, effort-based
understandings of success and, thus, the view that markets are just
assignors. Downward mobility tends to heighten economic insecurity,
dependence on public protection, and attachment to external attributions
of poverty, reducing support for market allocation. Meritocratic beliefs
are treated not as mediators but as moderators of these effects: among
strong meritocrats, upward mobility should amplify pro-commodification
attitudes, whereas downward mobility should attenuate -- or even reverse
-- its expected de-commodifying impact, as meritocratic interpretations
of effort lead individuals to view market outcomes as fair regardless of
direction. This inquiry contributes to research on market justice by
identifying the causal effects of mobility and using variation in
meritocratic beliefs to assess whether these effects follow a
meritocratic pathway.

This study focuses on Chile, an instructive case for examining
preferences for pension commodification. Despite sustained economic
growth, Chile combines high and persistent inequality with short-range
upward mobility and substantial barriers to higher-class positions
(\citeproc{ref-flores_top_2020}{Flores et al., 2020};
\citeproc{ref-salgado_uplifting_2025}{Salgado et al., 2025};
\citeproc{ref-torche_intergenerational_2014}{Torche, 2014}). It was the
first country worldwide in 1981 to fully replace a public pay-as-you-go
system with a mandatory, privately administered defined-contribution
scheme, while the state has progressively expanded a segmented
solidarity pillar to compensate for insufficient benefits
(\citeproc{ref-boccardo_30_2020}{Boccardo, 2020};
\citeproc{ref-madariaga_three_2020}{Madariaga, 2020};
\citeproc{ref-solimano_rise_2021}{Solimano, 2021}). In parallel---and
despite waves of protest from 2006 to 2019 against the private pension
system (\citeproc{ref-somma_no_2021}{Somma et al., 2021})---Chilean
subjectivities have been increasingly shaped by neoliberal discourses
and market logics, influencing attitudes toward pension distribution and
inequality (\citeproc{ref-araujo_desafios_2012}{Araujo \& Martuccelli,
2012}; \citeproc{ref-galvez_con_2023}{Gálvez, 2023}). The combination of
profound inequality, market-driven old-age security and contested
legitimacy makes Chile a critical case for analysing how mobility and
meritocratic beliefs structure support for pension commodification.

In this context, the questions that guide this research are as follows:

\begin{enumerate}
\def\labelenumi{(\arabic{enumi})}
\tightlist
\item
  How does intergenerational social mobility affect individuals'
  preferences for the commodification of pensions in Chile?
\item
  How do meritocratic beliefs condition or moderate this relationship?
\end{enumerate}

@@

\subsection{Preferences for pensions
commodification}\label{preferences-for-pensions-commodification}

Beyond the state's capacity to redistribute resources, market
institutions are central arenas for allocating socially valuable goods
and risks (\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}).
Markets are not mere aggregates of individual choices but social
institutions endowed with rules and normative meanings, built through
``moral projects, saturated with normativity'' that permeate everyday
thinking around fairness, effort and deservingness
(\citeproc{ref-fourcade_moral_2007}{Fourcade \& Healy, 2007};
\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019}). In this sense,
the economic order is mirrored in a moral economy: shared norms and
beliefs about what is considered a just distribution, embedded and
reinforced through institutions and policy feedback
(\citeproc{ref-svallforsMoralEconomyClass2006a}{Svallfors, 2006}). In an
era of privatisation and commodification, market logic has expanded into
core areas of social reproduction---childcare, healthcare, education and
pensions---stratifying access and quality
(\citeproc{ref-ferre_welfare_2023}{Ferre, 2023};
\citeproc{ref-gingrich_making_2011}{Gingrich, 2011}) and generating
feedback effects on public beliefs about welfare and market inequality
(\citeproc{ref-lindh_public_2015}{Lindh, 2015}). In this context,
support for market-based welfare provision has grown worldwide,
particularly among higher-income groups who view private alternatives as
more efficient or higher-quality
(\citeproc{ref-busemeyer_welfare_2020}{Busemeyer \& Iversen, 2020};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). These institutional
transformations provide the backdrop for analysing how far citizens
accept the commodification of social protection
(\citeproc{ref-satz_por_2019}{Satz, 2019}).

The legitimacy of market-based welfare is closely tied to beliefs about
distributive justice grounded in market principles. Beyond
stratification structures, citizens hold structured beliefs about how
resources ought to be allocated, reflecting causal attributions for
inequality, normative principles and expectations about deservingness
(\citeproc{ref-kluegel_beliefs_1981}{Kluegel \& Smith, 1981}). Research
on distributive justice examines conceptions of how goods and rewards
should be distributed and when inequality is considered just
(\citeproc{ref-castillo_legitimacy_2011}{Castillo, 2011};
\citeproc{ref-jasso_gender_1999}{Jasso \& Wegener, 1999}), while work on
redistributive preferences focuses on support for state-led mechanisms
that reduce inequality (\citeproc{ref-cavaille_fair_2025}{Cavaillé,
2025}). Situated at this intersection, the emerging literature on market
justice preferences conceptualises the market itself as a site of
redistribution and asks whether individuals regard market criteria as
fair bases for allocating wages and access to core welfare services
(\citeproc{ref-castillo_perceptions_2025}{Castillo, Laffert, et al.,
2025}; \citeproc{ref-lindh_public_2015}{Lindh, 2015};
\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}). In this
perspective, the expansion of market logic into welfare domains implies
that social services are treated as legitimate commodities that can be
priced and stratified (\citeproc{ref-busemeyer_welfare_2020}{Busemeyer
\& Iversen, 2020}), and empirical work suggests that such institutional
changes feed back into attitudes, reinforcing market-conforming
understandings of fairness and responsibility
(\citeproc{ref-zhu_policy_2015}{Ling Zhu \& Lipsmeyer, 2015}).

Market justice preferences refer to normative beliefs that legitimise
allocating goods and services according to market criteria such as
income and purchasing power
(\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al., 1999};
\citeproc{ref-lindh_public_2015}{Lindh, 2015}). Building on Lane's
(\citeproc{ref-lane_market_1986}{1986}) contrast between market and
political justice, market justice is a distributive principle that
endorses rewarding effort, productivity and skill rather than need and
equality as emphasised in welfare-state policies. In this study, I treat
market justice less as an abstract doctrine and more concretely as
preferences for commodification: support for treating welfare services
as commodities, so that better services follow from higher income or
ability to pay, and the resulting stratified access is seen as fair
(\citeproc{ref-lindh_public_2015}{Lindh, 2015}). Empirically, such
orientations are typically measured with items asking whether it is fair
that access to services depends on income, a strategy rooted in research
on justifications of capitalist inequality
(\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al., 1999}) and
applied to healthcare and education
(\citeproc{ref-castillo_changes_2025}{Castillo, Iturra, et al., 2025};
\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
\citeproc{ref-lee_fairness_2023}{Lee \& Stacey, 2023};
\citeproc{ref-vondemknesebeck_are_2016}{Von Dem Knesebeck et al.,
2016}). Moving beyond single domains, Lindh
(\citeproc{ref-lindh_public_2015}{2015}) develops a comparative index of
income-based access to healthcare and education, while Castillo,
Laffert, et al. (\citeproc{ref-castillo_perceptions_2025}{2025}) use the
same single-item measure but add the pensions domain in Chile. These
constructs gauge how individuals view market-generated inequalities as
legitimate and capture two core dimensions of market distribution: the
role of income in determining attainment and the framing of services as
commodities that can be bought and sold according to ability to pay
(\citeproc{ref-lindh_public_2015}{Lindh, 2015}).

This study examines support for the commodification of pensions, defined
as normative judgements on whether better pensions should depend on
income or individual contributions. I treat this as a domain-specific
form of market justice: pensions are conceived as commodities whose
allocation should follow contribution- and return-based criteria rather
than principles of need (\citeproc{ref-lane_market_1986}{Lane, 1986};
\citeproc{ref-mau_inequality_2015}{Mau, 2015}). The spread of funded,
privately managed schemes built on individual accounts has transformed
both the distributive and normative structure of old-age security
(\citeproc{ref-arenas_sistemas_2019}{Arenas, 2019};
\citeproc{ref-huber_political_2000}{Huber \& Stephens, 2000}).
Distributively, these schemes reassign risk from the collective to the
individual by tying benefits to formal employment, contribution density,
earnings and financial returns, making retirement income heavily
dependent on labour-market trajectories and capital-market success
(\citeproc{ref-madero-cabib_private_2019}{Madero-Cabib et al., 2019};
\citeproc{ref-oecd_pensions_2023}{OECD, 2023}). Normatively, they recast
pensions as the outcome of personal investment decisions rather than
collective obligations, promoting ideals of individual responsibility,
self-economisation and contribution-based desert
(\citeproc{ref-arza_pension_2008}{Arza, 2008};
\citeproc{ref-borzutzky_pension_2012}{Borzutzky, 2012}). Contributions
are increasingly framed as private assets rather than risk pooling, and
workers are expected to act as employee--entrepreneurs who choose
providers, monitor portfolios and ``make their money work'' over the
life course (\citeproc{ref-borzutzky_chiles_2016}{Borzutzky \& Hyde,
2016}; \citeproc{ref-kerner_pension_2020}{Kerner, 2020};
\citeproc{ref-mau_inequality_2015}{Mau, 2015}). As more people tie their
security expectations to private investments and financial markets,
their material interests become more deeply rooted in the market, and
their imperatives gain stronger normative support
(\citeproc{ref-mau_inequality_2015}{Mau, 2015}). When citizens are asked
whether higher pensions should be allocated to those who earn and
contribute more, they reflect support for the market allocation of a
critical social service and the legitimacy of the resulting risks and
inequalities in old age (\citeproc{ref-busemeyer_welfare_2020}{Busemeyer
\& Iversen, 2020}; \citeproc{ref-lindh_public_2015}{Lindh, 2015}).
